---
title: "STA 9750 · MP01 — Netflix Top 10"
author: "YASHPAL SAINI"
date: "October 02, 2025"
output:
  html_document:
    theme:
      version: 5
      bootswatch: flatly
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
    code_folding: hide
    df_print: paged
    css: styles.css
---

<section class="hero">
  <h1>Gourmet Cheeseburgers Across the Globe</h1>
  <p>We mine Netflix’s public Top 10 to highlight “gourmet cheeseburgers”—high-quality, mass-appeal hits—then summarize results as press releases for PR.</p>
  <div class="badges">
    <span class="badge">Quarto (.qmd)</span>
    <span class="badge">tidyverse</span>
    <span class="badge">DT</span>
    <span class="badge">ggplot2</span>
  </div>
</section>

::: callout
**Executive Summary.** We acquire the Netflix Tudum Top-10 data (global + per-country), fix “N/A” season labels, explore trends with tables & charts, then produce press-release-style highlights suitable for PR.
:::

# Introduction

The Netflix Top 10 provides a weekly pulse on what viewers watch worldwide. We answer the instructor tasks with **clear visuals and narrative**:
- How many countries are represented?
- Which programs dominate by **scale** (hours), **endurance** (weeks), and **breadth** (debut reach)?
- What PR-worthy stories emerge (e.g., *Stranger Things*)?

# Data Acquisition

```{r}
if (!require("tidyverse")) install.packages("tidyverse")
if (!require("DT")) install.packages("DT")
if (!require("lubridate")) install.packages("lubridate")
if (!require("stringr")) install.packages("stringr")
if (!require("scales")) install.packages("scales")

library(tidyverse)
library(readr); library(dplyr)
library(DT); library(lubridate); library(stringr); library(scales)

# Create data folder + download files if missing
if(!dir.exists(file.path("data","mp01"))){
  dir.create(file.path("data","mp01"), recursive = TRUE)
}

GLOBAL_TOP_10_FILENAME  <- file.path("data","mp01","global_top10_alltime.tsv")
COUNTRY_TOP_10_FILENAME <- file.path("data","mp01","country_top10_alltime.tsv")

if(!file.exists(GLOBAL_TOP_10_FILENAME)){
  download.file("https://www.netflix.com/tudum/top10/data/all-weeks-global.tsv",
                destfile = GLOBAL_TOP_10_FILENAME, mode = "wb")
}
if(!file.exists(COUNTRY_TOP_10_FILENAME)){
  download.file("https://www.netflix.com/tudum/top10/data/all-weeks-countries.tsv",
                destfile = COUNTRY_TOP_10_FILENAME, mode = "wb")
}

file.exists(GLOBAL_TOP_10_FILENAME) & file.exists(COUNTRY_TOP_10_FILENAME)
```

# Data Cleaning

We parse TSVs and convert literal `"N/A"` season labels into proper `NA`.

```{r}
GLOBAL_TOP_10 <- read_tsv(GLOBAL_TOP_10_FILENAME, show_col_types = FALSE) |>
  mutate(season_title = if_else(season_title == "N/A", NA_character_, season_title))

COUNTRY_TOP_10 <- read_tsv(
  COUNTRY_TOP_10_FILENAME,
  na = c("N/A"),
  show_col_types = FALSE
)

glimpse(GLOBAL_TOP_10, width = 80)
glimpse(COUNTRY_TOP_10, width = 80)
```

# KPIs at a Glance

```{r}
n_countries <- COUNTRY_TOP_10 |>
  distinct(country_name) |>
  filter(!is.na(country_name)) |>
  nrow()

# Top non-English film by cumulative weeks in global top 10
noneng_top <- GLOBAL_TOP_10 |>
  filter(str_detect(category, "Films"), !str_detect(category, "English")) |>
  group_by(show_title) |>
  summarise(max_weeks = max(cumulative_weeks_in_top_10, na.rm=TRUE), .groups="drop") |>
  arrange(desc(max_weeks)) |>
  slice(1)

# Longest film that ever appeared (convert hours->minutes)
add_runtime_minutes <- function(df){
  df |>
    mutate(runtime_minutes = if_else(!is.na(runtime), round(60*runtime), NA_real_))
}
longest_film <- GLOBAL_TOP_10 |>
  filter(str_detect(category, "Films")) |>
  add_runtime_minutes() |>
  filter(!is.na(runtime_minutes)) |>
  arrange(desc(runtime_minutes)) |>
  slice(1) |>
  select(show_title, runtime_minutes)

# Squid Game total global hours (all seasons)
squid_total <- GLOBAL_TOP_10 |>
  filter(str_detect(show_title, regex("Squid Game", ignore_case = TRUE)),
         str_detect(category, "TV")) |>
  summarise(total_hours = sum(weekly_hours_viewed, na.rm = TRUE))
```

<div class="kpis">
  <div class="card"><h3>Countries with Top-10 Data</h3><div class="big">`r scales::comma(n_countries)`</div><div class="tiny">Proxy for Netflix coverage</div></div>
  <div class="card"><h3>Top Non-English Film (by weeks)</h3><div class="big">`r noneng_top$show_title`</div><div class="tiny">`r noneng_top$max_weeks` cumulative weeks</div></div>
  <div class="card"><h3>Longest Film in Top-10</h3><div class="big">`r longest_film$show_title`</div><div class="tiny">`r longest_film$runtime_minutes` minutes</div></div>
  <div class="card"><h3>*Squid Game* Total Hours</h3><div class="big">`r scales::comma(squid_total$total_hours)`</div><div class="tiny">All seasons (global)</div></div>
</div>

# EDA

## Interactive preview (random 20 rows)

```{r}
GLOBAL_TOP_10 |>
  sample_n(size = min(20, nrow(GLOBAL_TOP_10))) |>
  mutate(`Runtime (Minutes)` = round(60*runtime)) |>
  select(week, category, weekly_rank, show_title, `Runtime (Minutes)`,
         weekly_hours_viewed, weekly_views, cumulative_weeks_in_top_10) |>
  rename(`Weekly Hours Viewed`=weekly_hours_viewed, `Weekly Views`=weekly_views,
         `Cumulative Weeks in Top 10`=cumulative_weeks_in_top_10,
         Week=week, Category=category, `Weekly Rank`=weekly_rank, Title=show_title) |>
  datatable(extensions="Buttons",
            options=list(dom="Bfrtip", buttons=c("copy","csv","excel"), pageLength=10))
```

## Visuals

### Category leaders by total hours
```{r}
most_hours_by_cat <- GLOBAL_TOP_10 |>
  group_by(category, show_title, season_title) |>
  summarise(total_hours = sum(weekly_hours_viewed, na.rm=TRUE), .groups="drop") |>
  group_by(category) |>
  slice_max(order_by = total_hours, n = 1, with_ties = FALSE)

library(ggplot2)
ggplot(most_hours_by_cat, aes(x=reorder(show_title, total_hours), y=total_hours, fill=category)) +
  geom_col() +
  coord_flip() +
  scale_y_continuous(labels = scales::label_number(scale_cut = scales::cut_short_scale())) +
  labs(x=NULL, y="Total Hours (global)", title="Category Leaders by Total Hours") +
  theme_minimal(base_size=13) +
  theme(legend.position="bottom")
```

### Top 10 programs by global hours (all time)
```{r}
top10_all <- GLOBAL_TOP_10 |>
  group_by(show_title) |>
  summarise(total_hours = sum(weekly_hours_viewed, na.rm=TRUE), .groups="drop") |>
  arrange(desc(total_hours)) |>
  slice(1:10)

ggplot(top10_all, aes(x=reorder(show_title, total_hours), y=total_hours)) +
  geom_col(fill="#5b8def") +
  coord_flip() +
  scale_y_continuous(labels = scales::label_number(scale_cut = scales::cut_short_scale())) +
  labs(x=NULL, y="Total Hours", title="Top 10 Programs by Global Hours (All Time)") +
  theme_minimal(base_size=13)
```

### *Stranger Things* — weekly hours (global)
```{r}
st_weekly <- GLOBAL_TOP_10 |>
  filter(str_detect(show_title, "Stranger Things")) |>
  group_by(week) |>
  summarise(hours = sum(weekly_hours_viewed, na.rm=TRUE), .groups="drop")

ggplot(st_weekly, aes(x=week, y=hours)) +
  geom_line(linewidth=0.9, color="#00c2ff") +
  scale_y_continuous(labels = scales::label_number(scale_cut = scales::cut_short_scale())) +
  labs(x=NULL, y="Weekly Hours", title="Stranger Things — Weekly Hours (Global)") +
  theme_minimal(base_size=13)
```

# Exploratory Questions (with Narrative)

## Q1. How many different countries does Netflix operate in (proxy)?
**Answer:** **`r n_countries` countries.**  
**Context.** Distinct country names in the per-country dataset proxy Netflix’s footprint, since weekly charts are only published where service is active and data meets reporting standards.

## Q2. Which non-English film spent the most cumulative weeks in the global Top 10?
**Answer:** **`r noneng_top$show_title`** with **`r noneng_top$max_weeks` weeks**.  
**Why it matters.** Extended global runs show stories resonate across languages. Titles like this are PR gold: durable visibility, long tail discovery, and catalogue lift.

```{r}
# Context distribution: how exceptional is the winner?
noneng_weeks <- GLOBAL_TOP_10 |>
  filter(str_detect(category,"Films"), !str_detect(category,"English")) |>
  group_by(show_title) |>
  summarise(max_weeks = max(cumulative_weeks_in_top_10, na.rm=TRUE), .groups="drop")
summary_noneng <- noneng_weeks |>
  summarise(median_weeks = median(max_weeks, na.rm=TRUE),
            p90 = quantile(max_weeks, .90, na.rm=TRUE))
summary_noneng
```

## Q3. What is the longest film to appear in the global Top 10? How long (minutes)?
**Answer:** **`r longest_film$show_title`** — **`r longest_film$runtime_minutes` minutes**.  
**Context.** While longer runtimes can challenge completions, strong propositions overcome frictions; this outlier shows audiences invest when the value is clear.

```{r}
film_runtime_stats <- GLOBAL_TOP_10 |>
  filter(str_detect(category,"Films")) |>
  add_runtime_minutes() |>
  summarise(median_min = median(runtime_minutes, na.rm=TRUE),
            p75 = quantile(runtime_minutes, .75, na.rm=TRUE))
film_runtime_stats
```

## Q4. For each category, which program has the most total global hours?
See the **Category leaders** chart.  
**Why it matters.** These are tentpoles—good for franchise building, renewals, and timed PR.

```{r}
most_hours_by_cat |>
  mutate(total_hours_fmt = scales::comma(total_hours)) |>
  select(category, show_title, season_title, total_hours_fmt) |>
  datatable(options = list(pageLength=5, searching=FALSE, info=FALSE),
            rownames = FALSE)
```

## Q5. Which TV show had the longest consecutive run in a country’s Top 10? Where and how long?
```{r}
is_tv <- str_detect(COUNTRY_TOP_10$category, "TV")
ctv <- COUNTRY_TOP_10[is_tv, ] |>
  arrange(country_name, show_title, season_title, week) |>
  group_by(country_name, show_title, season_title) |>
  mutate(
    week = as.Date(week),
    gap = as.integer(week - lag(week)),
    new_run = if_else(is.na(gap) | gap > 7, 1L, 0L),
    run_id = cumsum(replace_na(new_run, 0L))
  ) |>
  ungroup()

run_spans <- ctv |>
  group_by(country_name, show_title, season_title, run_id) |>
  summarise(start = min(week), end = max(week),
            weeks = as.integer((end - start) / 7) + 1L, .groups="drop") |>
  arrange(desc(weeks)) |>
  slice(1)
run_spans
```
**Answer:** **`r run_spans$show_title`** — **`r run_spans$weeks` weeks** in **`r run_spans$country_name`** (from **`r run_spans$start`** to **`r run_spans$end`**).  
**Why it matters.** Long, unbroken runs indicate cultural resonance and word-of-mouth—perfect for local PR amplification.

## Q6. Country with truncated service history (less than ~200 weeks). When did it end?
```{r}
coverage <- COUNTRY_TOP_10 |>
  group_by(country_name) |>
  summarise(n_weeks = n_distinct(week),
            last_week = max(week), .groups="drop") |>
  arrange(n_weeks)
truncated <- coverage |> slice(1)
truncated
```
**Answer:** **`r truncated$country_name`**, last reported week **`r truncated$last_week`**.  
**Note.** Truncated histories often reflect exits or reporting changes.

## Q7. Total global hours for *Squid Game* (all seasons)
**Answer:** **`r scales::comma(squid_total$total_hours)` hours**.  
**Context.** The scale validates franchise status and continued catalog draw.

## Q8. *Red Notice* views in 2021 (approx., using hours ÷ runtime)
```{r}
rn_runtime_hours <- 1 + 58/60
rn_2021 <- GLOBAL_TOP_10 |>
  filter(show_title == "Red Notice", lubridate::year(week) == 2021) |>
  summarise(total_hours = sum(weekly_hours_viewed, na.rm=TRUE))
rn_views_2021 <- rn_2021$total_hours / rn_runtime_hours
rn_2021$total_hours; rn_views_2021
```
**Answer:** ≈ **`r scales::comma(round(rn_views_2021))` views** in 2021.  
**Method.** Total hours ÷ hours-per-view is a reasonable approximation.

## Q9. Films that hit #1 in the US after debuting below #1 (+ most recent)
```{r}
us_films <- COUNTRY_TOP_10 |>
  filter(country_name == "United States", str_detect(category, "Films")) |>
  group_by(show_title) |>
  summarise(first_rank = min(weekly_rank, na.rm=TRUE),
            ever_num1  = any(weekly_rank == 1, na.rm=TRUE),
            last_week  = max(week),
            .groups="drop")

risers <- us_films |>
  filter(ever_num1, first_rank > 1) |>
  arrange(desc(last_week))
head(risers, 10)
```
**Answer:** **`r nrow(risers)` films**; most recent climber: **`r risers$show_title[1]`** (last week **`r risers$last_week[1]`**).  
**Interpretation.** Late surges often reflect strong word-of-mouth and PR acceleration.

## Q10. TV show/season that charted in the most countries in its *debut* week
```{r}
country_debuts <- COUNTRY_TOP_10 |>
  filter(str_detect(category, "TV")) |>
  group_by(show_title, season_title, country_name) |>
  summarise(debut_week = min(week), .groups="drop")

spread <- COUNTRY_TOP_10 |>
  inner_join(country_debuts, by=c("show_title","season_title","country_name")) |>
  filter(week == debut_week) |>
  group_by(show_title, season_title, week) |>
  summarise(n_countries = n_distinct(country_name), .groups="drop") |>
  arrange(desc(n_countries)) |>
  slice(1)
spread
```
**Answer:** **`r spread$show_title`** (**`r spread$season_title`**) — **`r spread$n_countries` countries** on debut (week of **`r spread$week`**).  
**Why it matters.** Wide day-one reach supports coordinated global launches and cross-market creatives.

# Press Releases (PR-Style)

::: {.card}
### Stranger Things Season 5 Set to Captivate a Global Fanbase
**Los Gatos, CA — `r format(Sys.Date(), "%B %d, %Y")`.** Across its first four seasons, *Stranger Things* amassed **`r scales::comma(sum(GLOBAL_TOP_10 |> filter(str_detect(show_title, "Stranger Things")) |> pull(weekly_hours_viewed), na.rm=TRUE))` hours** globally and maintained durable Top-10 presence across **`r n_countries` markets**. With multi-season momentum and worldwide appeal, Season 5 is poised for another blockbuster debut.
:::

::: {.card}
### Non-English Films Keep Winning Audiences Worldwide
**Amsterdam — `r format(Sys.Date(), "%B %d, %Y")`.** Non-English features continue to show extended global legs. The top performer (**`r noneng_top$show_title`**) held the global Top 10 for **`r noneng_top$max_weeks` weeks**, far above typical runs (see distribution above), highlighting the universal resonance of local stories.
:::

::: {.card}
### English-Language Films Show Broad Global Momentum
**London — `r format(Sys.Date(), "%B %d, %Y")`.** Category leaders (see charts) underscore robust appetite for English-language tentpoles with global reach. The breadth of engagement supports coordinated cross-market campaigns and talent spotlights.
:::

# Conclusion & Next Steps

- The Top-10 data reveals programs with exceptional **endurance** (long runs), **scale** (global hours), and **breadth** (debut reach).  
- These metrics guide PR narratives (global beats, local spotlights) and catalogue positioning.  
- **Next:** Slice by region and language, then align assets (trailers, key art) per priority market to maximize impact.
