---
title: "STA 9750 · MP01 — Netflix Top 10"
author: "YASHPAL SAINI"
date: "October 02, 2025"
output:
  html_document:
    theme:
      version: 5
      bootswatch: flatly
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
    code_folding: hide
    df_print: paged
    css: styles.css
---

<section class="hero">
  <h1>Gourmet Cheeseburgers Across the Globe</h1>
  <p>High‑quality content that wins mass‑market audiences. We mine Netflix’s public Top 10 to identify global “gourmet cheeseburgers”.</p>
  <div class="badges">
    <span class="badge">Reproducible</span><span class="badge">KPI Cards</span><span class="badge">Interactive Tables</span><span class="badge">Beautiful Charts</span>
  </div>
</section>

<div class="callout">
<b>Executive Summary.</b> We acquire the Netflix Tudum Top 10 data (global + per‑country), fix minor quirks (e.g., <code>"N/A"</code> season labels), and conduct an EDA to spotlight programs with durable, multinational appeal. Results are summarized as press releases for PR usage.
</div>

# Introduction

The Netflix Top 10 gives a weekly pulse of what viewers are watching worldwide. This report answers the instructor’s tasks with clear visuals and editorial context:
- What breadth of countries does the data cover?
- Which programs sustain long runs or achieve exceptional total hours?
- How do iconic titles (e.g., <i>Stranger Things</i>) trend over time?
- Where are the PR‑worthy stories?

# Data Acquisition

**Source**: Netflix Tudum weekly Top 10 datasets (global and per‑country).

```{r}
if (!require("tidyverse")) install.packages("tidyverse")
if (!require("DT")) install.packages("DT")
if (!require("lubridate")) install.packages("lubridate")
if (!require("stringr")) install.packages("stringr")
if (!require("scales")) install.packages("scales")

library(tidyverse)
library(readr); library(dplyr)
library(DT); library(lubridate); library(stringr); library(scales)

# Create data folder and download files if needed
if(!dir.exists(file.path("data", "mp01"))){ dir.create(file.path("data", "mp01"), showWarnings=FALSE, recursive=TRUE) }
GLOBAL_TOP_10_FILENAME <- file.path("data", "mp01", "global_top10_alltime.tsv")
if(!file.exists(GLOBAL_TOP_10_FILENAME)){
  download.file("https://www.netflix.com/tudum/top10/data/all-weeks-global.tsv",
                destfile=GLOBAL_TOP_10_FILENAME, mode="wb")
}
COUNTRY_TOP_10_FILENAME <- file.path("data", "mp01", "country_top10_alltime.tsv")
if(!file.exists(COUNTRY_TOP_10_FILENAME)){
  download.file("https://www.netflix.com/tudum/top10/data/all-weeks-countries.tsv",
                destfile=COUNTRY_TOP_10_FILENAME, mode="wb")
}
file.exists(GLOBAL_TOP_10_FILENAME) & file.exists(COUNTRY_TOP_10_FILENAME)
```

# Data Cleaning

Two changes get us analysis‑ready: (1) parse TSVs; (2) change literal <code>"N/A"</code> to proper <code>NA</code> in <code>season_title</code>.

```{r}
GLOBAL_TOP_10 <- read_tsv(GLOBAL_TOP_10_FILENAME, show_col_types = FALSE) |>
  mutate(season_title = if_else(season_title == "N/A", NA_character_, season_title))

COUNTRY_TOP_10 <- read_tsv(COUNTRY_TOP_10_FILENAME, na = c("N/A"), show_col_types = FALSE)
```

## Structure checks

```{r}
glimpse(GLOBAL_TOP_10, width = 80)
glimpse(COUNTRY_TOP_10, width = 80)
```

# KPIs at a Glance

```{r}
n_countries <- COUNTRY_TOP_10 |> distinct(country_name) |> filter(!is.na(country_name)) |> nrow()

noneng_top <- GLOBAL_TOP_10 |>
  filter(str_detect(category, "Films"), !str_detect(category, "English")) |>
  group_by(show_title) |>
  summarise(max_weeks = max(cumulative_weeks_in_top_10, na.rm=TRUE), .groups="drop") |>
  arrange(desc(max_weeks)) |>
  slice(1)

add_runtime_minutes <- function(df){ df |> mutate(runtime_minutes = if_else(!is.na(runtime), round(60*runtime), NA_real_)) }
longest_film <- GLOBAL_TOP_10 |>
  filter(str_detect(category, "Films")) |>
  add_runtime_minutes() |>
  filter(!is.na(runtime_minutes)) |>
  arrange(desc(runtime_minutes)) |>
  slice(1) |>
  select(show_title, runtime_minutes)

squid_total <- GLOBAL_TOP_10 |>
  filter(str_detect(show_title, regex("Squid Game", ignore_case=TRUE)), str_detect(category,"TV")) |>
  summarise(total_hours = sum(weekly_hours_viewed, na.rm=TRUE))
```

<div class="kpis">
  <div class="card"><h3>Countries with Top‑10 Data</h3><div class="big">`r scales::comma(n_countries)`</div><div class="tiny">Coverage proxy for Netflix presence</div></div>
  <div class="card"><h3>Top Non‑English Film (by weeks)</h3><div class="big">`r noneng_top$show_title`</div><div class="tiny">`r noneng_top$max_weeks` cumulative weeks</div></div>
  <div class="card"><h3>Longest Film in Top‑10</h3><div class="big">`r longest_film$show_title`</div><div class="tiny">`r longest_film$runtime_minutes` minutes</div></div>
  <div class="card"><h3>*Squid Game* Total Hours</h3><div class="big">`r scales::comma(squid_total$total_hours)`</div><div class="tiny">All seasons (global)</div></div>
</div>

# EDA

## Interactive preview (20 random rows)

```{r echo=FALSE, message=FALSE, warning=FALSE}
GLOBAL_TOP_10 |>
  sample_n(size = min(20, nrow(GLOBAL_TOP_10))) |>
  mutate(`Runtime (Minutes)` = round(60*runtime)) |>
  select(week, category, weekly_rank, show_title, `Runtime (Minutes)`, weekly_hours_viewed, weekly_views, cumulative_weeks_in_top_10) |>
  rename(`Weekly Hours Viewed`=weekly_hours_viewed, `Weekly Views`=weekly_views, `Cumulative Weeks in Top 10`=cumulative_weeks_in_top_10, Week=week, Category=category, `Weekly Rank`=weekly_rank, Title=show_title) |>
  datatable(extensions="Buttons",
            options=list(dom="Bfrtip", buttons=c("copy","csv","excel"), pageLength=10))
```

## Visuals

### Category leaders by total hours

```{r echo=FALSE}
most_hours_by_cat <- GLOBAL_TOP_10 |>
  group_by(category, show_title, season_title) |>
  summarise(total_hours = sum(weekly_hours_viewed, na.rm=TRUE), .groups="drop") |>
  group_by(category) |>
  slice_max(order_by=total_hours, n=1, with_ties=FALSE)

ggplot(most_hours_by_cat, aes(x=reorder(show_title, total_hours), y=total_hours, fill=category)) +
  geom_col() +
  coord_flip() +
  scale_y_continuous(labels=scales::label_number(scale_cut = scales::cut_short_scale())) +
  labs(x=NULL, y="Total Hours (global)", title="Category Leaders by Total Hours") +
  theme_minimal(base_size=13) +
  theme(legend.position="bottom")
```

### Top 10 programs by global hours (all time)

```{r echo=FALSE}
top10_all <- GLOBAL_TOP_10 |>
  group_by(show_title) |>
  summarise(total_hours = sum(weekly_hours_viewed, na.rm=TRUE), .groups="drop") |>
  arrange(desc(total_hours)) |>
  slice(1:10)

ggplot(top10_all, aes(x=reorder(show_title, total_hours), y=total_hours)) +
  geom_col(fill="#5b8def") +
  coord_flip() +
  scale_y_continuous(labels=scales::label_number(scale_cut = scales::cut_short_scale())) +
  labs(x=NULL, y="Total Hours", title="Top 10 Programs by Global Hours (All Time)") +
  theme_minimal(base_size=13)
```

### Stranger Things — weekly hours (global)

```{r echo=FALSE}
st_weekly <- GLOBAL_TOP_10 |>
  filter(str_detect(show_title, "Stranger Things")) |>
  group_by(week) |>
  summarise(hours = sum(weekly_hours_viewed, na.rm=TRUE), .groups="drop")

ggplot(st_weekly, aes(x=week, y=hours)) +
  geom_line(linewidth=0.9, color="#00c2ff") +
  scale_y_continuous(labels=scales::label_number(scale_cut = scales::cut_short_scale())) +
  labs(x=NULL, y="Weekly Hours", title="Stranger Things — Weekly Hours (Global)") +
  theme_minimal(base_size=13)
```

# Exploratory Questions (with Narrative)

## Q1. How many countries does Netflix operate in (proxy from viewing data)?

**Answer:** **`r n_countries` countries.**  
**Context.** The per‑country Top 10 dataset contains distinct country names where weekly charts are reported. While not an official list of operating markets, it is a strong proxy for Netflix’s global footprint because charts are only published where service is active and data quality is sufficient.

## Q2. Which non‑English‑language film spent the most cumulative weeks in the global Top 10?

**Answer:** **`r noneng_top$show_title`** with **`r noneng_top$max_weeks` weeks** in the global Top 10.  
**Why it matters.** Long global runs for non‑English titles show that audiences readily cross language barriers when the story travels. That sustained visibility supports marketing beats, late‑window discovery, and catalog uplift.

```{r}
# Extra context: how typical is this run among non-English films?
noneng_weeks <- GLOBAL_TOP_10 |>
  filter(str_detect(category,"Films"), !str_detect(category,"English")) |>
  group_by(show_title) |>
  summarise(max_weeks = max(cumulative_weeks_in_top_10, na.rm=TRUE), .groups="drop")

summary_noneng <- noneng_weeks |>
  summarise(median_weeks = median(max_weeks, na.rm=TRUE), p90 = quantile(max_weeks, .90, na.rm=TRUE))

summary_noneng
```
*Reading guide.* The winner’s run substantially exceeds the median non‑English film (median = **`r summary_noneng$median_weeks`** weeks; 90th percentile ≈ **`r summary_noneng$p90`** weeks), reinforcing its exceptional staying power.

## Q3. What is the longest film (English or non‑English) to appear in the global Top 10, and how long is it?

**Answer:** **`r longest_film$show_title`** with a runtime of **`r longest_film$runtime_minutes` minutes**.  
**Context.** Runtime can influence completion and repeat viewing. This title stands above typical Top‑10 film lengths; it suggests audiences will invest in longer features when the proposition is compelling.

```{r}
film_runtime_stats <- GLOBAL_TOP_10 |>
  filter(str_detect(category,"Films")) |>
  add_runtime_minutes() |>
  summarise(median_min = median(runtime_minutes, na.rm=TRUE),
            p75 = quantile(runtime_minutes, .75, na.rm=TRUE))
film_runtime_stats
```
On average, Top‑10 film runtimes cluster around **`r film_runtime_stats$median_min`** minutes (median), with a heavier tail up to **`r film_runtime_stats$p75`** minutes (75th percentile).

## Q4. For each category, which program has the most total hours of global viewership?

See *Category leaders* chart above.  
**Context.** We aggregate weekly hours by `(category, title, season)` and pick the maximum per category. These leaders serve as anchors for future programming, tentpole promotions, and franchise strategy.

```{r echo=FALSE}
most_hours_by_cat |>
  mutate(total_hours_fmt = scales::comma(total_hours)) |>
  select(category, show_title, season_title, total_hours_fmt) |>
  datatable(options = list(pageLength=5, searching=FALSE, info=FALSE), rownames=FALSE)
```

## Q5. Which TV show had the longest consecutive run in a country’s Top 10? How long and where?

```{r}
is_tv <- str_detect(COUNTRY_TOP_10$category, "TV")
ctv <- COUNTRY_TOP_10[is_tv, ] |>
  arrange(country_name, show_title, season_title, week) |>
  group_by(country_name, show_title, season_title) |>
  mutate(
    week = as.Date(week),
    gap = as.integer(week - lag(week)),
    new_run = if_else(is.na(gap) | gap > 7, 1L, 0L),
    run_id = cumsum(replace_na(new_run, 0L))
  ) |>
  ungroup()

run_spans <- ctv |>
  group_by(country_name, show_title, season_title, run_id) |>
  summarise(start = min(week), end = max(week), weeks = as.integer((end - start)/7) + 1L, .groups="drop") |>
  arrange(desc(weeks)) |>
  slice(1)
```
**Answer:** **`r run_spans$show_title`** held a **`r run_spans$weeks`‑week** consecutive run in **`r run_spans$country_name`** (from **`r run_spans$start`** to **`r run_spans$end`**).  
**Why it matters.** Long uninterrupted runs in a market indicate strong word‑of‑mouth and cultural resonance, helpful for regional PR campaigns.

## Q6. Netflix provides over 200 weeks of service history for all but one country. Which country is it and when did service cease?

```{r}
coverage <- COUNTRY_TOP_10 |>
  group_by(country_name) |>
  summarise(n_weeks = n_distinct(week), last_week = max(week), .groups="drop") |>
  arrange(n_weeks)

truncated <- coverage |> slice(1)
```
**Answer:** **`r truncated$country_name`**, with final reporting week **`r truncated$last_week`**.  
**Note.** A truncated history often reflects a service exit or a pause in reporting.

## Q7. What is the total viewership of the TV show *Squid Game* (all seasons, hours)?

**Answer:** **`r scales::comma(squid_total$total_hours)` hours** globally across all seasons.  
**Context.** This consolidates weekly hours across seasons and weeks at a global level. The magnitude underscores the show’s event‑level status.

## Q8. *Red Notice* (runtime 1h58m): approximately how many views did it receive in 2021?

```{r}
rn_runtime_hours <- 1 + 58/60
rn_2021 <- GLOBAL_TOP_10 |>
  filter(show_title == "Red Notice", lubridate::year(week) == 2021) |>
  summarise(total_hours = sum(weekly_hours_viewed, na.rm=TRUE))
rn_views_2021 <- rn_2021$total_hours / rn_runtime_hours
```
**Answer:** About **`r scales::comma(round(rn_views_2021))` views** in 2021.  
**Method.** Total 2021 hours divided by runtime (in hours) gives an approximate “complete‑view” count.

## Q9. How many Films reached #1 in the US after debuting below #1? Most recent example?

```{r}
us_films <- COUNTRY_TOP_10 |>
  filter(country_name == "United States", str_detect(category, "Films")) |>
  group_by(show_title) |>
  summarise(first_rank = min(weekly_rank, na.rm=TRUE),
            ever_num1  = any(weekly_rank == 1, na.rm=TRUE),
            last_week  = max(week), .groups="drop")

risers <- us_films |> filter(ever_num1, first_rank > 1) |> arrange(desc(last_week))
```
**Answer:** **`r nrow(risers)` films** achieved a climb to #1 after a lower debut; most recent climber: **`r risers$show_title[1]`** (last chart week **`r risers$last_week[1]`**).  
**Interpretation.** Climbing to #1 post‑launch signals growing momentum, often driven by social chatter, PR, or recommendation loops.

## Q10. Which TV show/season charted in the most countries in its debut week?

```{r}
country_debuts <- COUNTRY_TOP_10 |>
  filter(str_detect(category, "TV")) |>
  group_by(show_title, season_title, country_name) |>
  summarise(debut_week = min(week), .groups="drop")

spread <- COUNTRY_TOP_10 |>
  inner_join(country_debuts, by=c("show_title","season_title","country_name")) |>
  filter(week == debut_week) |>
  group_by(show_title, season_title, week) |>
  summarise(n_countries = n_distinct(country_name), .groups="drop") |>
  arrange(desc(n_countries)) |>
  slice(1)
```
**Answer:** **`r spread$show_title`** (**`r spread$season_title`**) debuted in **`r spread$n_countries` countries** in the week of **`r spread$week`**.  
**Why it matters.** Wide day‑one reach indicates a truly global hook and helps justify global launches and cross‑market creative.

# Press Releases (for PR Team)

<div class="card">
<h3>Stranger Things Season 5 Set to Captivate a Global Fanbase</h3>
<p><b>Los Gatos, CA — `r format(Sys.Date(), "%B %d, %Y")`.</b> Across its first four seasons, <i>Stranger Things</i> amassed <b>`r scales::comma(sum(GLOBAL_TOP_10 |> filter(str_detect(show_title, "Stranger Things")) |> pull(weekly_hours_viewed), na.rm=TRUE))` hours</b> globally and maintained a durable presence in weekly Top 10 charts across <b>`r n_countries` markets</b>. With multi‑season momentum and worldwide appeal, Season 5 is poised for another blockbuster debut.</p>
</div>

<div class="card">
<h3>Non‑English Films Keep Winning Audiences Worldwide</h3>
<p><b>Amsterdam — `r format(Sys.Date(), "%B %d, %Y")`.</b> Non‑English features continue to show extended global legs. The top performer (<b>`r noneng_top$show_title`</b>) held the global Top 10 for <b>`r noneng_top$max_weeks` weeks</b>, far above the typical run (median <b>`r summary_noneng$median_weeks`</b>). The sustained performance highlights the universal resonance of local stories.</p>
</div>

<div class="card">
<h3>English‑Language Films Show Broad Global Momentum</h3>
<p><b>London — `r format(Sys.Date(), "%B %d, %Y")`.</b> Category leaders (see charts) underscore robust appetite for English‑language tentpoles with global reach. The breadth of engagement supports coordinated cross‑market campaigns and talent spotlights.</p>
</div>

# Conclusion & Next Steps

- The Top 10 data reveals programs with exceptional <b>endurance</b> (long consecutive runs), <b>scale</b> (total global hours), and <b>breadth</b> (wide debut reach).  
- These metrics guide PR narratives (global press beats, local spotlights) and catalogue positioning.  
- <b>Next:</b> Slice leaders by region and language, and pair with title‑level creative (trailers, key art) in each priority market for maximized impact.

<div class="callout note"><b>Limitations.</b> Top 10 is weekly and may exclude borderline titles; runtimes for older entries can be missing; and view “counts” derived from hours ÷ runtime are approximations.</div>

<footer style="margin-top:24px;color:#667085;font-size:.9rem">© `r format(Sys.Date(), "%Y")` — Analysis by Yash Saini · Built with R Markdown, tidyverse, DT, ggplot2.</footer>
